%No one could have ever imagined how the last year would turn out to be. Essential keywords include quaranteene, isolation and social distancing. Consequently, I believe this puts bits and pieces more easily in perspective since we are able to spend more time with ourself and our own thoughts. Therefore, I truly have many persons to thank for pulling me through this thesis.

The dream of having a social life during my masters was utterly shattered by the ongoing covid-19 pandemic. I recall joking ``see you after easter'' when the University locked down two weeks before easter the year $2020$. Now, $60$ weeks later, I'm still waiting for the reunion.
I was not certain that I would come out on the other side with a degree at all. For that, I truly have many persons to thank for pulling me through.

I'd like to thank my main supervisor, Morten, for being such an inspiring role model. I did not fancy studying until I encountered your positive energy and pure pleasure of research.


Secondly, I would like to thank my supervisors Marianne, Øyvind, Sebastian and Lasse for excellent counseling through both fun and difficult times. Thank you for shining light upon my path and letting me know when my good ideas were bad and bad ideas were good (mostly just bad).

% For this, I owe a thanks to the great people of MENA and computational science group.

Thirdly, I would like to thank the people that made the years of studying the best so far. Thank you, Mohamed, Jens, Erik, Jørn, Andreas, and the rest of the people at MENA and CS for sharing obligs, discussions, early mornings, late evenings, lunch breaks, pump-sundays at Athletica, hangovers, cabin trips, and occasional nonsense.

I would also like to thank my family's moral support and encouragement even though I am positive that they had no idea what I was talking about.


%Pre-corona, I had no idea that the years of attending the University of Oslo would be the best years I've had so far. In particular, I've shared many obligs, discussions, early mornings, late evenings, pump-sundays at Athletica, hangovers, cabin trips, and occasional nonsense with Mohamed, Jens, Erik, Jørn, Andreas, and the rest of the people at MENA.

Last but not least, I would like to thank my girlfriend Hanna for limitless support throughout this journey and for sacrificing a dining table so I could have a place to study. I may have written a thesis, but I believe you have \emph{suffered} a thesis.
